Viscous Transport in Eroding Porous Media

### 2019-10-05 docs/transport/erosion_transport.tex
working on edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{Bryan is here}
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here, the streamline originates at $(x_0,y_0)$ on the inlet and its
arclength is calculated once it passes the parallel outlet $x = x_0 +
d$.  The hydraulic tortuosity is defined by taking the average over all
points on the cross-section
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0}
  {\displaystyle\int_{S}u_1(x_0,y_0)\,dy_0},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is the inlet cross-section $x = x_0$ that is perpendicular to
the $x$-axis, $u_1(y_0)$ is the $x$ component of the velocity at the
initial point of the streamline, and $d$ is still the distance between
the inlet and outlet. Therefore, the tortuosity satisfies $T \geq 1$,
and $T=1$ only if no grains are present.

The tortuosity is defined in terms of the length of streamlines between
a starting and ending point. The starting point is typically along a
cross-section $S$, but there are several options for the ending
point~\cite{dud-koz-mat2011}.  The choice adopted here is to use
streamlines that originate at $S$ and terminate at a parallel
cross-section.  We use the cross-sections $x=-1$ and $x=1$ that are at
the start and end of the porous region.  This choice of averaging allows
us to also compute the tortuosity in terms of an area integral.
Assuming that the flow is incompressible and not re-entrant, meaning
that all streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent to~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_1(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\cite{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in Section~\ref{sec:results}.

There have been efforts to relate the tortuosity to the porosity.  For
example, Matyka et al.~\cite{mat-kha-koz2008} propose the models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $\phi$ is the porosity and $p>0$ is a fitting parameter.  In
Section~\ref{sec:results}, we compare these four models with the
tortuosity of eroding porous geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes between
neighboring grains. Two grains are said to be neighbors if they share an
edge of a Delanuay configuration that uses nodes placed at the center of
each eroding grain. (b) The distribution of the pore sizes of an eroded
geometry. The black curve is the Weibull distribution with the same
first two moments as the data.}
\end{figure}

While transport in a porous media depends on the porosity, it also
depends on the placement of the grains.  In particular, the grain
placement affects velocity scales~\cite{ali-par-wei-bre2017},
correlation structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
more. In addition, the grain configuration is important for pore network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} that rely on
descriptions of the pore throat sizes. To characterize the grain
placement, we compute distributions of pore sizes between neighboring
grains.  To define neighboring grains, we form a Delaunay triangulation
using nodes placed at the center of each eroding grain and at a
collection of points around the boundary of the porous region.  Then, we
say that two grains are neighbors if their centers share an edge of the
triangulation~\cite{dea-qua-bir-jua2018}. The pores for a eroded
geometry are illustrated in Figure~\ref{fig:Eroding100gap}(a). We do not
consider pores between eroding bodies and the solid wall $\Gamma$, so
some of the grains near the porous region boundary only have two
neighbors. Having defined the pore sizes, we form its distribution and
compare it with the Weibull distribution (black curve) in
Figure~\ref{fig:Eroding100gap}(b) . A more complete statistical analysis
of the pore size distribution is in Section~\ref{sec:Eroding100}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{sec:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In line with our previous work~\cite{qua-moo2018}, we use two meshes to
simulate erosion. The integral equation is solved by discretizing the
boundary of the geometry at a set of collocation points distributed
equally in arclength (Section~\ref{sec:spatialDiscretization}).
Depending on the proximity of the target point to the source points, the
quadrature rule is either the trapezoid rule or the Barycentric
quadrature rule (Section~\ref{sec:bary}).  The selection algorithm that
determines which quadrature method is applied is described in
Section~\ref{sec:fmm}.  Once the shear stress is computed, the bodies
are eroded a single time step by using a {\thL}
discretization~\cite{hou-low-she1994, moore2013self}.  The time stepping methods for
erosion and passive particles are described in Section~\ref{sec:time}

@@

In Figure~\ref{fig:Eroding100gap_hist}, we plot
histograms of the pore pore sizes at six porosities during the erosion
process. These types of distributions have been described with Weibull
distributions~\cite{ioa-cha1993}, and we superimpose the Weibull
distribution with the same first two moments as the data.  The
parameters of the distribution, $(k,\lambda)$, are included in the
caption of Figure~\ref{fig:Eroding100gap_hist}.  In
Figure~\ref{fig:Eroding100gap_mean_var}, we plot the mean and variance
of the pore sizes as a function of the porosity.  Interestingly, for
porosities less than $\phi = 85\%$, the mean pore size grows linearly
and the variance remains nearly flat. Since a channelized geometry has
large variance, this indicates that channelization is less prevalent at
low porosities.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here, the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\cite{dud-koz-mat2011}.  The
hydraulic tortuosity is defined by taking the average over all points on
the inlet cross-section
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0}
  {\displaystyle\int_{S}u_1(x_0,y_0)\,dy_0},
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.

The definition we made for the tortuosity allows us to compute the
tortuosity in terms of an area integral.  Assuming that the flow is
incompressible and not re-entrant, meaning that all streamlines connect
the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent to~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_1(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\cite{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in Section~\ref{sec:results}.

There have been efforts to relate the tortuosity, $T$, to the porosity,
$\phi$.  For example, Matyka et al.~\cite{mat-kha-koz2008} propose the
models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $p>0$ is a fitting parameter.  In Section~\ref{sec:results}, we
compare these four models with the tortuosity of eroding porous
geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\cite{ali-par-wei-bre2017}, correlation
structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019,berhanu2012shape}, and pore
network models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}. To
characterize the grain placement, we compute distributions of pore sizes
between neighboring grains.  To define neighboring grains, we form a
Delaunay triangulation using nodes placed at the center of each eroding
grain and at a collection of points around the boundary of the porous
media.  Then, we say that two grains are neighbors if their centers
share an edge of the triangulation~\cite{dea-qua-bir-jua2018}. The pores
of an eroded geometry are illustrated in
Figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
Figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\cite{ioa-cha1993}.  In Section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.
\begin{figure}[H]
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes between
neighboring grains. Two grains are said to be neighbors if they share an
edge of a Delanuay triangulation with nodes at the center of each
eroding grain. (b) The distribution of the pore sizes of an eroded
geometry. The black curve is the Weibull distribution with the same
first two moments as the data.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{sec:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{Bryan is here}
In line with our previous work~\cite{qua-moo2018}, we use two meshes to
simulate erosion. The integral equation is solved by discretizing the
boundary of the geometry at a set of collocation points distributed
equally in arclength (Section~\ref{sec:spatialDiscretization}).
Depending on the proximity of the target point to the source points, the
quadrature rule is either the trapezoid rule or the Barycentric
quadrature rule (Section~\ref{sec:bary}).  The selection algorithm that
determines which quadrature method is applied is described in
Section~\ref{sec:fmm}.  Once the shear stress is computed, the bodies
are eroded a single time step by using a {\thL}
discretization~\cite{hou-low-she1994, moore2013self}.  The time stepping methods for
erosion and passive particles are described in Section~\ref{sec:time}

@@

HERE
In Figure~\ref{fig:Eroding100gap_hist}, we plot
histograms of the pore pore sizes at six porosities during the erosion
process. These types of distributions have been described with Weibull
distributions~\cite{ioa-cha1993}, and we superimpose the Weibull
distribution with the same first two moments as the data.  The
parameters of the distribution, $(k,\lambda)$, are included in the
caption of Figure~\ref{fig:Eroding100gap_hist}.  In
Figure~\ref{fig:Eroding100gap_mean_var}, we plot the mean and variance
of the pore sizes as a function of the porosity.  Interestingly, for
porosities less than $\phi = 85\%$, the mean pore size grows linearly
and the variance remains nearly flat. Since a channelized geometry has
large variance, this indicates that channelization is less prevalent at
low porosities.
